NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-03-09 main.tex
Updates from Overleaf

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access is still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. Recent advances using machine learning approaches utilizing publicly available databanks, for example to solve protein structures~\cite{jumper21}, emphasize the increasing importance of such resources. 
%The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

@@

While the NMRlipids Databank currently contains only lipid bilayer systems, it can readily be expanded to contain other molecules, such as disordered proteins or sugars. The powerful combination of the overlay databank structure and a community-wide open collaboration can be applied also when building databases in other fields, particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist. Therefore, the approach introduced here opens up possibilities to create databases enabling data-driven and machine learning applications in new fields.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The new NMRlipids databank will support the development in both of these directions. The automatic quality evaluation and ranking of simulations against experimental data in the NMRlipids databank can be used to select the best parameters for simulations of large biomolecular complexes and to guide optimizations of new force field parameters for both atomistic and coarse grained simulations. These issues are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amount of MD simulation via NMRlipids databank-API enables wide range of novel data-driven analyses and machine learning applications. For example, large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enables analyses of rare phenomena beyond current possibilities for a single research group as demonstrated for cholesterol flip--flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}). This will also pave the way of applications of MD simulations in new fields as demonstrated here by analysing the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}) that is essential parameter in pharmacokinetic modeling and MRI imaging~\cite{nitsche19,topgaard20}. Furthermore, the NMRlipids databank-API paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties using NMRlipids databank as a training set. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict proteins structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientist who do not typically have access to large scale MD simulation data.
%and for analyses of rare phenomena that require exceptional amounts of simulation data.
%On the other hand, universal naming convention enables systematic comparisons between simulations having different model parameters and thereby different naming conventions.
%
%The NMRlipids databank will be of great benefit to scientists involved in MD simulations. They will be able to rapidly evaluate the quality of MD simulations in order to filter out potentially misleading results or facilitate force field parameter development~\cite{antila22b}. The selection of best models for a specific application is demonstrated for PC/PE lipid mixtures in Fig.~\ref{fig:quality}, and previously for PC/PS lipid mixtures~\cite{antila22b}, and lipid headgroups~\cite{bacle21}. 
%
Different types of expected applications of the NMRlipids databank in wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the NMRlipids databank.

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip--flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientist in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for POPC lipids (Fig.~\ref{fig:quality}), headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:POPCPOPEapls}) containing membranes. &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids databank, extracting continuum parameters for membranes. &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning type of models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules.  
    \end{tabular}
    \caption{Examples on applications of the NMRlipids databank.}
    \label{tab:applications}
\end{table}

The main practical limitations in building open access databanks of molecular dynamics simulations have been the required commitment in long term support for hardware and software maintenance, and the lack of incentives for researchers to share the data. The NMRlipids databank circumvents these challenges with the overlay databank design and open collaboration approach developed in the NMRlipids project~\cite{botan15}. In this model, the file storage is distributed to publicly available stable repositories (\textit{Data layer} in Fig.~\ref{fig:overlay} A) and maintenance of the databank does not depend on individual scientists or groups because all its version controlled content is available with an open access licence (\textit{Databank layer} in Fig.~\ref{fig:overlay} A). Incentive to share the data is created in the NMRlipids open collaboration by offering authorship in published articles to the contributors following the NMRlipids project protocol~\cite{botan15}. The current NMRlipids databank contains only lipid bilayer simulations, but the concept can also be applied to other biomolecules, such as disordered proteins and membrane-protein systems, or other fields where similar barriers to establish publicly accessible databanks exist.

>>>>

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access are still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. Recent advances using machine learning approaches utilizing publicly available databanks, for example to solve protein structures~\cite{jumper21}, emphasize the increasing importance of such resources. 
%The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

@@

While the NMRlipids Databank currently contains only lipid bilayer systems, its key elements that enable the programmatic access to large scale MD simulation data can be applied also for other molecules, such as disordered proteins or sugars. The powerful combination of the overlay databank structure and a community-wide open collaboration can be used to build databases that enable data-driven and machine learning applications also in other fields where storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist, such as assignment of NMR spectra~\cite{klukowski22}.

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids databank will support the development in both of these directions. The automatic quality evaluation and ranking of simulations against experimental data can be used to select the best parameters for simulations of large biomolecular complexes and to guide optimizations of new force field parameters for both atomistic and coarse grained simulations. The accuracy of simulations is becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amount of MD simulation data via NMRlipids databank-API enables wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip--flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields as demonstrated here by analysing the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}) that is essential parameter in pharmacokinetic modeling and MRI imaging~\cite{nitsche19,topgaard20}. Furthermore, the NMRlipids databank-API delivers access to a training set data that paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict proteins structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientist who do not typically have access to large scale MD simulation data.
%and for analyses of rare phenomena that require exceptional amounts of simulation data.
%On the other hand, universal naming convention enables systematic comparisons between simulations having different model parameters and thereby different naming conventions.
%
%The NMRlipids databank will be of great benefit to scientists involved in MD simulations. They will be able to rapidly evaluate the quality of MD simulations in order to filter out potentially misleading results or facilitate force field parameter development~\cite{antila22b}. The selection of best models for a specific application is demonstrated for PC/PE lipid mixtures in Fig.~\ref{fig:quality}, and previously for PC/PS lipid mixtures~\cite{antila22b}, and lipid headgroups~\cite{bacle21}. 
%
Different types of expected applications of the NMRlipids databank in wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the NMRlipids databank.

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip--flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientist in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:POPCPOPEapls}) containing membranes. &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for the development of coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids databank, extracting continuum parameters for membranes. &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning type of models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules.  
    \end{tabular}
    \caption{Examples on applications of the NMRlipids databank.}
    \label{tab:applications}
\end{table}

Potential benefits of sharing MD simulation data are widely recognized by different stakeholders from scientists~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} to funders and journal staff, but data sharing with efficient tools for upcycling have been rare. The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. The first issue is solved in the NMRlipids databank using the overlay design where raw data is distributed to already publicly available decentralized locations while the core of the databank is composed only of metadata stored in a version controlled git repository with an open access licence. On the other hand, the open collaboration approach developed in the NMRlipids project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such approach are demonstrated here for membrane simulations, but the concept can be applied also not only to other biomolecules, such as disordered proteins and membrane-protein systems, but also in other fields where similar barriers hinder the machine learning revolution, such as assignment of NMR spectra~\cite{klukowski22}.
